Sample Length at Which Block Bootstrapping is Effective for Estimation of Parameter of Time Series

### 2023-01-07 Manuscript/manuscript.tex
New Code

\title{Time Series Length at Which Block Bootstrapping is Effective for Estimation of Parameter and Its Variance}

@@

\begin{abstract}
Block bootstrapping is a method that can be used for estimating a parameter of a time
series. It involves splitting a series into blocks (in order to account for the time
factor) and re-sampling the blocks to create many new bootstrapped time series.
This method becomes more effective as the length of the time series increases. 
The question for this study is how does one determine at what length the block
bootstrap method stops being effective to estimate a parameter of a time
series.

@@

Block bootstrapping is a widely used method in Statistics. The concept for block 
bootstrapping was developed independently by \citet{hall1985resampling}, \citet{carlstein1986use}, and 
\citet{kunsch1989jackknife}. \citet{radovanov2014comparison} It has been applied to a variety of 
different domains such 
as econometrics \citep{mackinnon2006bootstrap} and meteorology \citep{varga2017generalised}. It can be used for 
situations in which there is temporal dependence within a population, and the goal is the estimation or testing a hypothesis about a parameter. Assuming that the sample is infinitely 
large, the method will work perfectly. However, for a finite sample size, the method will 
not work as well as expected. \citet{hesterberg2015teachers} notes that while intervals based on the percentiles from nonparametric bootstrapping are more accurate than t-intervals for larger sample sizes, they are less accurate for smaller sample sizes. In the context of planning an applied 
statistical procedure, for which a large sample size is not always practical, it may be helpful to know which sample size is necessary for the method to work at an acceptable level. 
The goal of this study is to find a threshold or range of sample sizes at which block bootstrap 
is an effective method for estimating a parameter and its variance. This is true if 1 - $\alpha$/2 \% (for some predetermined $\alpha$) of confidence intervals created from samples of the same size recover the parameter.

\citet{nevitt2001performance} find that a sample size of 200-1000 is usually sufficient for estimation using the bootstrap method (simple resampling with no blocks involved). However, block bootstrapping is a different situation altogether because the sample is split into blocks.

\citet{goncalves2005bootstrap} found that standard error estimates from block bootstrapping small 
samples 
may be significantly more accurate than inference from closed-form asymptotic estimates. That study focused on the estimation of a coefficient parameter within the context of linear regression. Still, the block bootstrap percentile confidence intervals under-covered the parameter even for n = 1024. This study aims to find how small the sample size can be for time series (which has a time dependence factor which can be compared to a coefficient of a linear regression) in order for the coverage rate to be close to the confidence level.

For this study, many block bootstrap simulations were conducted with RStudio. At the base
level, we are block-bootstrapping an auto-regressive process with true mean 0, and some AR(1) coefficient.
Bootstrapping is the term for creating new samples of the same size by re-sampling from
the original sample with replacement. Basic bootstrapping works well for samples
that are do not have a dependence factor. However, for a time series such as an auto-regressive process,
a different procedure is required to account for the time dependence. In such a case,
the series is split into blocks that can overlap (moving block bootstrap) or not(non-moving block bootstrap). These blocks are then re-sampled to create a new
bootstrapped time series sample. Point estimates of a parameter from many block bootstrapped samples are used to create a distribution to create an interval estimation of the parameter.

In this experiment, the mean, auto-correlation function (with a maximum lag of 1), and variance of a 1000 block-bootstrapped time series samples are computed, 
and a 95\% confidence intervals of these parameters is created. This is replicated 4000 times, 
and the proportion of confidence intervals that recover the true known parameter 
(coverage rate) is recorded. The goal of this is to observe how successful the bootstrapping process
is at estimating a parameter and its variance. The key variable being observed was, 
the length of the time series, or the size of the sample. It is known that as n
increases, block bootstrapping will become a less accurate method for estimation
(the coverage rates will decrease). The question is at what range of sample lengths does this
start to become a problem and what range of sample lengths is necessary to reach a coverage rate of 95\%.

In this experiment, there are certain factors that are expected to affect the coverage
rates. As the Auto-Regressive (AR) coefficient (the time dependence) of the time series 
increases,
we expect the coverage rates of the confidence intervals to decrease.
The coverage rates are also affected by the size of the blocks - more specifically,
how the size of the blocks (l) relates to the size of the time series (n) - 
and whether or not they overlap. It is known that as the size of the time series 
increases, the optimal block length should increase, and the ratio of the block length to 
the time series length should decrease. l = n$^{1/3}$ is typically considered the optimal 
block length according to asymptotic theory. \citep{buhlmann1999block} Simulations were conducted using this optimal function of l = n$^{1/3}$.

In order to solve the question of what n is necessary for effective block bootstrapping,
while still accounting for all these factors, the simulations were repeated for
variations of the AR coefficient, target parameter, and non-moving vs moving method. For each of these combinations, block bootstrap results were recorded for a wide range of sample sizes.

@@

Block bootstrapping is a method that can be applied to a time series (or any sample of data for that matter) to estimate a parameter and its variance. Suppose that a sample of size n of a time series is given. In regular bootstrap procedure, the new sample of the same size as the original would just be created by resampling observations with replacement. In the case of a time series, in order to account for the temporal dependence, the time series can be split into blocks, typically of the same size. The block should be of size l large enough to include time dependence, yet small enough to include some variance. As n increases, an ideal l should also increase, but the ratio of l to n should decrease. To achieve this, l is often assigned a value as a function of n. A common function that is considered the best by much previous literature is l = n$^{1/3}$. However not all n sizes are perfect squares, and l must be a whole number, so the ceiling of this function must be used. A bootstrapped sample is created by taking a sample of n / l blocks with replacement to create a new series of size n. The designer of the study can choose whether the blocks overlap or not. If the blocks can never overlap, it is called a non-moving block bootstrap. In this case, the original sample is split evenly into n / l blocks, and the same number of blocks is sampled without replacement to create a bootstrapped sample. If the blocks do overlap, it is called a moving block bootstrap. n / l samples are taken randomly from the original sample, but they do not have to be from a set of evenly spaced blocks. An estimate of the parameter is computed from the bootstrapped sample. This procedure can be repeated many times (typically 1000) to create a sampling distribution of the estimator. Using this distribution of estimates, a 1 - $\alpha$/2 \% confidence interval for the parameter can be created using the $\alpha$/2 and 1 - $\alpha$/2 percentiles of sample estimates.

>>>>

\title{Sample Length at Which Block Bootstrapping is Effective for Estimation of Parameter of Time Series}

@@

\begin{abstract}
Block bootstrapping is a method that can be used for estimating a parameter of a time
series. It involves splitting a series into blocks (in order to account for the temporal dependence
factor) and re-sampling the blocks to create many new bootstrapped pseudo-samples of the time series.
This method becomes more effective as the length of the time series increases. 
The question that this study seeks to answer is what sample length is necessary for the block
bootstrap method to be effective in estimating a parameter of a time
series.

@@

Block bootstrapping is a widely used method in Statistics. The concept for block 
bootstrapping was developed independently by \citet{hall1985resampling}, \citet{carlstein1986use}, and 
\citet{kunsch1989jackknife}. \citet{radovanov2014comparison} It has been applied to a variety of 
different domains such 
as econometrics \citep{mackinnon2006bootstrap} and meteorology \citep{varga2017generalised}. It can be used for 
situations in which there is temporal dependence within a population, and the goal is the estimation or testing a hypothesis about a parameter. Assuming that the sample is infinitely 
large, the method will work perfectly. However, for a finite sample size, the method will 
not work as well as expected. \citet{hesterberg2015teachers} notes that while percentile-based confidence intervals from nonparametric bootstrapping are more accurate than t-intervals for larger sample sizes, they are less accurate for smaller sample sizes. In the context of planning an applied 
statistical procedure, for which a large sample size is not always practical, it may be helpful to know how large the sample must be for the method to work at an acceptable level. 
The goal of this study is to find a threshold or range of sample sizes at which block bootstrap 
is an effective method for estimating a parameter. The method is effective if approximately 1 - $\alpha$/2 \% (for some predetermined significance level $\alpha$) of confidence intervals created from samples of the same size recover the parameter.

\citet{nevitt2001performance} find that a sample size of 200-1000 is usually sufficient for estimation using the bootstrap method (simple resampling with no blocks involved). However, in this study, the focus is block bootstrapping a temporally dependent sample.This is a different situation altogether, first, because the sample has a dependence factor that must be accounted for, and second because block bootstrapping involves splitting the sample into blocks.

\citet{goncalves2005bootstrap} found that standard error estimates from block bootstrapping small 
samples 
may be significantly more accurate than inference from closed-form asymptotic estimates. That study focused on the estimation of a coefficient parameter within the context of linear regression. Still, the block bootstrap percentile confidence intervals under-covered the parameter even for n = 1024. This study aims to find how small the sample size can be for a time series (which has a time dependence factor comparable to a coefficient of a linear regression) in order for the coverage rate to be close to the confidence level.

For this study, the block bootstrap method was applied to time series simulations in R.

%At the base level, we are block-bootstrapping an auto-regressive process with true mean 0, and some AR(1) coefficient.

Bootstrapping is the procedure of creating new pseudo-samples of the same size by resampling observations with replacement from
the original sample. Basic bootstrapping works well for samples
that are do not have a dependence factor. However, for a time series such as an auto-regressive or moving-average process,
a different procedure is required to account for the time dependence. In such a situation, instead of resampling observations, blocks of the original sample can be resampled (these blocks must be of a certain length by which the sample length is divisible) to create a pseudo-sample. Point estimates of a parameter calculated from many block bootstrapped pseudo-samples are then used to create a distribution to create an interval estimation of the parameter.

%In this experiment, the mean, auto-correlation function (with a maximum lag of 1), and variance of a 1000 block-bootstrapped time series samples are computed, and 95\% confidence intervals of these parameters is created.

The various ways which these intervals can be constructed are described in \citet{diciccio1996bootstrap} and \citet{rice2006mathematical}. They will be discussed later in a review of the block bootstrap method. For the purposes of this study (in which the true parameters are known), in order to establish whether or not the method is effective, the block bootstrapping of a simulated sample can be replicated many times, 
and the proportion of confidence intervals that recover the true known parameter 
(coverage rate) is recorded. The goal of this is simply to observe how successful the block bootstrapping procedure
is at estimating a parameter. The key variable being observed was, the sample length (n). It is known that as n
increases, block bootstrapping should become a more accurate and precise method for estimation
(the coverage rates will approach the theoretical confidence level and the interval widths will decrease). The question is what is the minimum sample length necessary to reach a coverage rate of 1 - $\alpha$/2 \%.

In this experiment, there are certain variables that are expected to affect the performance of block bootstrap. As the the time dependence (AR or MA coefficient) of the time series 
increases, we expect the coverage rates of the confidence intervals to decrease.
The coverage rates are also affected by the size of the blocks - more specifically,
how the size of the blocks (l) relates to the size of the time series (n) - 
and whether or not they overlap. It is known that as the size of the time series 
increases, the optimal block length should increase, and the ratio of the block length to 
the time series length should decrease. l = n$^{1/3}$ is typically considered the optimal 
block length according to asymptotic theory. \citep{buhlmann1999block} Therefore, simulations in this study were conducted using this optimal function of l = n$^{1/3}$.

%In order to solve the question of what sample length is necessary for effective block bootstrapping while still accounting for all these factors, the simulations were repeated for different variations of the MA coefficient, target parameter, and non-moving vs moving method.

For each of these variations, results for block bootstrap estimation of mean, MA coefficient, and standard deviation were recorded for a wide range of sample length.

@@

Block bootstrapping is a method that can be applied to a non-stationary process to estimate a parameter. Suppose that a sample of length n of a time series is given. In regular bootstrap procedure, the new pseudo-sample of the same size as the original sample would be created by simply resampling observations from the with replacement. In the case of a time series, in order to account for the temporal dependence, the time series can be split into blocks, typically of the same size. The block should be of size l large enough to include time dependence, yet small enough to include some variance. l should also divide n. As n increases, an ideal l should also increase, but the ratio of l to n should decrease. To achieve this, l is often assigned a value as a function of n. A common function that is considered the best by much previous literature is l = n$^{1/3}$. To make things sample for this study, only perfect cubes were used for the sample length. A bootstrapped sample is created by taking a sample of n / l blocks with replacement to create a new pseudo-sample of size n. The designer of the study can choose whether the blocks overlap or not. If the blocks can never overlap, it is called a non-moving block bootstrap. In this case, the original sample is split evenly into n / l blocks, and the same number of blocks are sampled with replacement to create a bootstrapped pseudo-sample. If the blocks do overlap, it is called a moving block bootstrap. n / l samples are taken randomly from the original sample, but they do not have to be from a set of evenly spaced blocks. An estimate of the parameter (or parameters) is computed from the bootstrapped sample. This procedure can be repeated many times to create a sampling distribution of the estimator. In this study, 1000 bootstrapped pseudo-samples were created for each simulation. Using this distribution of point estimates, a 1 - $\alpha$/2 \% confidence interval for the parameter can be created. The most simplistic method for constructing this interval is the percentile confidence interval. For a parameter $\theta$, the percentile confidence interval takes the form: \[ [ ]\]. Another method for constructing this interval is the empirical confidence interval, which takes the form: \[ [ ]\]. The last method used in this study is the bias-corrected and accelerated 
(BCA) confidence interval, which takes the form: \[ [ ]\].
